---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---

```{r libraries, echo=TRUE, message=FALSE}
library(lubridate)
library(dplyr)
library(ggplot2)

## Format display numbers to display without scientific notation
options(scipen=999)
```

## Loading and preprocessing the data
```{r loading,echo=TRUE}

rawDataFile <- "repdata-data-activity.zip"
downloadURL <- "https://github.com/psiu/RepData_PeerAssessment1/raw/master/repdata-data-activity.zip"

if(!file.exists(rawDataFile)){
    download.file(downloadURL, rawDataFile)
}

## Load data from data file
activity <- read.csv(unz(rawDataFile, "activity.csv"))

## Format dates
activity$date <- ymd(activity$date)

activity <- tbl_df(activity)
```


## What is mean total number of steps taken per day?
```{r mean_steps,echo=TRUE}
# Aggregate number of steps per day
daily_steps <- activity %>% group_by(date) %>% summarize(steps = sum(steps))

mean_steps <- mean(daily_steps$steps, na.rm=TRUE)
median_steps <- median(daily_steps$steps, na.rm=TRUE)

hist(daily_steps$steps, col = "green", xlab = "Total number of steps per day", ylab = "Frequency (Number of days)", main ="Historgram on frequency of steps per day")
abline(v = mean_steps, col = "red")
text(mean_steps + 2000, 20 , "Mean")
```

Looking at the raw data set:

The mean number of steps per day is `r round(mean_steps,2)`, denoted by the red line above

The median number of steps per day is `r median_steps`


## What is the average daily activity pattern?
```{r mean_daily_activity,echo=TRUE}
interval_avg_steps <- activity %>% group_by(interval) %>% summarize(avg_steps = mean(steps, na.rm=TRUE))

high_score <- max(interval_avg_steps$avg_steps)

ggplot(interval_avg_steps, aes(x= interval, y = avg_steps)) +
    geom_line() +
    xlab("Interval") +
    ylab("Steps (average)") +
    ggtitle("Average steps over daily time interval")
```

The maximum number of steps in any given interval period is `r round(high_score, 2)`

## Imputing missing values
```{r na_count,echo=TRUE}
num_NA <- sum(is.na(activity$steps))
```

Number of NA values in the data step is `r num_NA`

In an attempt to normalize the data, we will replace the null values in the data set with the average value within the same interval.

```{r imput,echo=TRUE}

## Copy data set
activity_norm <- activity
## Replace NA values with imput strategy, using the interval_avg_steps as a lookup table for average interval step value
for(i in 1:nrow(interval_avg_steps)) {
  look <- as.numeric(interval_avg_steps[i,1])
  replacement <- as.numeric(interval_avg_steps[i,2])
  activity_norm[activity_norm$interval == look & is.na(activity_norm$steps),]$steps <- replacement
}

##############
## Evaluate new data set

# Aggregate number of steps per day
daily_steps_norm <- activity_norm %>% group_by(date) %>% summarize(steps = sum(steps))

mean_steps <- mean(daily_steps_norm$steps, na.rm=TRUE)
median_steps <- median(daily_steps_norm$steps, na.rm=TRUE)

hist(daily_steps_norm$steps, col = "green", xlab = "Total number of steps per day", ylab = "Frequency (Number of days)", main ="Historgram on frequency of steps per day (normalized)")
abline(v = mean_steps, col = "red")
text(mean_steps + 2000, 20 , "Mean")
```

Normalized data set:

The mean number of steps per day is `r round(mean_steps,2)`, denoted by the red line above

The median number of steps per day is `r round(median_steps,2)`

Imputing the missing values did not have a big impact on the estimated total daily number of steps.


## Are there differences in activity patterns between weekdays and weekends?
```{r weekday_weekends,echo=TRUE}

## Add weekday / weekend factor
activity_norm$week <- "Weekday"
activity_norm[wday(activity_norm$date) == 1 | wday(activity_norm$date) == 7, "week"] <- "Weekend"
activity_norm$week <- as.factor(activity_norm$week)

## Average the data over interval and weekday/weekend
averaged_norm <- activity_norm %>% group_by(week, interval) %>% summarize(steps = mean(steps))

ggplot(averaged_norm, aes(x = interval, y = steps)) +
    facet_grid(week ~ .) +
    geom_line() +
    labs(title = "Average steps on weekdays vs weekend", x = "Interval (min)", y = "Steps (average)")

```